---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Cynthia Deng"
date: "02/19/2026"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r}
staff <- read_csv("data/instructional-staff.csv")
```

### Exercise 1

```{r}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```
```{r}
staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()
```
```{r}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Percentage of Faculty Types Hired by Year",
       x= "Year",
       y= "Percentage of Hires",
       color = "Faculty Type")

```

### Exercise 2

To better communicate that part-time faculty have increased over time, the plot can visually emphasize the part-time faculty line while de-emphasizing the others. For example, making the part-time line thicker and using a bold color, and fading the remaining faculty types into light gray.
```{r}
staff_long %>%
  ggplot(aes(
    x = as.integer(year),
    y = value,
    group = faculty_type
  )) +

  geom_line(
    data = subset(staff_long, faculty_type != "Part-Time Faculty"),
    color = "gray",
    linewidth = 0.8
  ) +

  geom_line(
    data = subset(staff_long, faculty_type == "Part-Time Faculty"),
    color = "purple",
    linewidth = 1.5
  ) +
  labs(
    title = "Growth of Part-Time Faculty Over Time",
    x = "Year",
    y = "Percentage of instructional staff"
  ) +
  theme_minimal()

```

### Exercise 3

```{r}
fisheries <- read_csv("data/fisheries.csv")
```
The original visualization does not communicate a clear story because plotting every country at once makes the visualization cluttered and difficult to interpret. Instead of trying to include hundreds of categories, it makes more sense to focus on the top-producing countries so the main patterns are visible. Additionally, a line plot would not be appropriate because the data do not represent change over time, but rather a single-year comparison across countries. A stacked bar chart is a better option because it clearly shows each country’s total production while also highlighting how much comes from capture versus aquaculture.

```{r}
fish_long <- fisheries %>%
  pivot_longer(
    cols = c(capture, aquaculture),
    names_to = "type",
    values_to = "tons"
  )

# Find top 15 countries by total production
top_countries <- fish_long %>%
  group_by(country) %>%
  summarise(total = sum(tons)) %>%
  slice_max(total, n = 15)

fish_top <- fish_long %>%
  filter(country %in% top_countries$country)

# Stacked bar chart
ggplot(fish_top,
       aes(x = reorder(country, tons, sum),
           y = tons,
           fill = type)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Fishery Production by Country (2016)",
    subtitle = "Top 15 producers, split by Capture vs Aquaculture",
    x = "Country",
    y = "Production (tons)",
    fill = "Production type"
  ) +
  theme_minimal()
```
### Stretch Practice
```{r}
data("Whickham")

```
```{r}
data("Whickham")
?Whickham
library(performance)

```
1. These data come from an observational study, not an experiment. The researchers did not assign people to be smokers or non-smokers. Instead, they simply recorded each woman’s existing smoking status at the baseline survey (1972–1974) and then followed up 20 years later to see whether she was alive or dead.

2.The dataset contains 1314 observations. Each observation represents one woman who participated in the Whickham survey.

3. There are 3 variables: outcome, smoker, and age

```{r}
ggplot(Whickham, aes(x = outcome)) +
  geom_bar() +
  labs(
    title = "Survival Outcome After 20 Years",
    x = "Outcome",
    y = "Count"
  ) +
  theme_minimal()

```
```{r}
ggplot(Whickham, aes(x = smoker)) +
  geom_bar() +
  labs(
    title = "Smoking Status at Baseline",
    x = "Smoker",
    y = "Count"
  ) +
  theme_minimal()

```
```{r}
ggplot(Whickham, aes(x = age)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Age Distribution of Participants",
    x = "Age (years)",
    y = "Frequency"
  ) +
  theme_minimal()

```

4. Smoking status should be associated with higher mortality rates. In other words, women who were smokers at baseline would be more likely to have the outcome Dead, while women who were non-smokers would be more likely to have the outcome Alive.

5. 
```{r}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(proportion = n / sum(n))

ggplot(Whickham, aes(x = smoker, fill = outcome)) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportion of Survival Outcomes by Smoking Status",
    x = "Smoking Status at Baseline",
    y = "Proportion of Participants",
    fill = "Outcome After 20 Years"
  ) +
  theme_minimal()

```
This stacked bar chart shows the conditional proportions of survival outcomes (Alive vs Dead) within each smoking group. Overall, most women in both groups were still alive after 20 years. Surprisingly, the proportion of women who died is slightly higher among non-smokers than smokers. Specifically, about 32% of non-smokers are in the “Dead” category compared to about 24% of smokers.

6.
```{r}
Whickham <- Whickham %>%
  mutate(
    age_cat = case_when(
      age <= 44 ~ "18-44",
      age > 44 & age <= 64 ~ "45-64",
      age > 64 ~ "65+"
    )
  )

```
7. 
```{r}
Whickham %>%
  count(smoker, age_cat, outcome)
```
```{r}
ggplot(Whickham, aes(x = smoker, fill = outcome)) +
  geom_bar(position = "fill") +
  facet_wrap(~ age_cat) +
  labs(
    title = "Survival Outcomes by Smoking Status, Faceted by Age Group",
    x = "Smoking Status at Baseline",
    y = "Proportion of Participants",
    fill = "Outcome After 20 Years"
  ) +
  theme_minimal()

```

When we facet the survival outcome plot by age category, the relationship between smoking and mortality changes substantially. In the youngest group (18–44), nearly all participants survive regardless of smoking status, although smokers show a slightly higher proportion of deaths. In the middle age group (45–64), the difference becomes more noticeable: smokers have a higher conditional probability of being dead after 20 years compared to non-smokers. In the oldest group (65+), mortality is very high for both smokers and non-smokers, but smokers again show a slightly greater proportion of deaths.When we compare smokers and non-smokers within the same age groups, the expected pattern emerges: smoking is generally linked to worse survival outcomes.